A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Shameless self-reference

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
    | [] -> None
    | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
  let rec quicksort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
\end{minted}
While \ocaml{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

A better approach for reasoning about program equivalences comes from
\emph{structured recursion schemes}, and their associated algebraic laws.  In
particular, \ocaml{quicksort} is an instance of a \emph{divide-and-conquer}
algorithm which, in the functional programming literature is known as a
hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
that every recursive program can be turned into an hylomorphism, which implies
that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
program equivalences, since they can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to (semi-)automatic
parallelisations~\cite{Gibbons96:Third}.

@@

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{quicksort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function
  | [] -> []
  | h :: t ->
    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
    let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
    app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}

>>>>

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
    | [] -> None
    | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
  let rec qsort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> qsort smaller @ (x::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

A better approach for reasoning about program equivalences comes from
\emph{structured recursion schemes}, and their associated algebraic laws.  In
particular, \ocaml{qsort} is an instance of a \emph{divide-and-conquer}
algorithm which, in the functional programming literature is known as a
hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
that every recursive program can be turned into an hylomorphism, which implies
that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
program equivalences, since they can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to (semi-)automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.

@@

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function
  | [] -> []
  | h :: t ->
    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
    let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
    app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
